\bf Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-15 manuscript/main.tex
improved writing on 4.1.1 truncation

	\subsubsection{Finite-basis approximation of the predictors and regression coefficients}\label{sec: finite basis approximation}	
	Given a twice-differentiable basis $\{b_l(t)\}$ of $L^2([0,1])$, the functional predictors and the corresponding regression coefficients can be decomposed into
	\begin{equation*}
		X_{ij}(t) = \sum_{l=1}^\infty \theta_{ijl} \, b_l(t), \quad 
		\beta_j(t) = \sum_{l=1}^\infty \eta_{0,jl} \, b_l(t),
	\end{equation*}
	where $\theta_{ijl}$ and $\eta_{0,jl}$ are the coefficients of $X_{ij}(t)$ and $\beta_j(t)$ corresponding to the $l$th basis function $b_l(t)$, respectively.
Since functional predictors are observed at finitely many discrete time points, we approximate them by truncating the expansion at $M$ terms, yielding a finite-dimensional projection. A moderately large $M$ (e.g., 15 or 20) suffices to capture rich functional variation, and no fine-tuning is needed, as our method controlls smoothness  via penalization (Section~\ref{sec: Regularization of PLS Components}). We denote the truncated space by
	\begin{equation}
		\mathcal{B} := \operatorname{span} \bigl( b_1(t), \ldots, b_M(t) \bigr).
	\end{equation}
	 We denote the  predictors projected onto $\mathcal{B}$ and the resulting coefficients as
	\begin{equation*}
		\widetilde{X}_{ij}(t) = \sum_{l=1}^M \theta_{ijl} \, b_l(t), \quad 
		\tilde{\beta}_j(t) = \sum_{l=1}^M \eta_{0,jl} \, b_l(t).
	\end{equation*}
	The $i$th  hybrid predictor with projected functional parts, is a tuple, denoted as
	\begin{equation*}
		\widetilde{W}_i := 
		(
		X_{i1}, \ldots, X_{iK}, \boldsymbol{Z}_i
		).
	\end{equation*}
	We adopt the following notation for the coefficients:
	\begin{equation}
		\boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top \in \mathbb{R}^M,
		\quad
		\boldsymbol{\eta}_{0j} := (\eta_{0,j1}, \ldots, \eta_{0,jM})^\top  \in \mathbb{R}^M,
	\end{equation}
	and
	\begin{equation}
		\Theta_j := (\boldsymbol{\theta}_{1j}, \ldots, \boldsymbol{\theta}_{nj})^\top  \in \mathbb{R}^{n \times M}.
	\end{equation}

The observations in the hybrid PLS problem can be compactly represented by the tuple
\begin{equation}\label{def:problem_instance}
	(B^\ast, \Theta, \mathbf{y}),
\end{equation}
defined as follows.
Let 
$B, B^{''} \in \mathbb{R}^{M \times M}$
denote the Gram matrices of the basis functions and their second derivatives, respectively, with entries  
\begin{equation}
\end{equation}
\[
B_{m, m'} := \langle b_m, b_{m'} \rangle = \int_0^1 b_m(t)\, b_{m'}(t)\, dt,
\]
for $m, m' = 1, \ldots, M$.
Then define $B^\ast := \operatorname{blkdiag}(B, \ldots, B, I_p) \in \mathbb{R}^{(MK + p) \times (MK + p)}$,  
$\Theta := (\Theta_1, \ldots, \Theta_K, Z) \in \mathbb{R}^{n \times (MK + p)}$,  
and $\mathbf{y} := (y_1, \ldots, y_n)^\top$.
	\begin{remark}
While it is possible to use a different basis for each functional predictor, we adopt a common basis for all predictors to simplify the notation. However, the definition of $B^\ast$ remains general enough to accommodate different bases across the $K$ predictors.
\end{remark}

@@

	\begin{proposition}\label{proposition:eigen_noregul}
Let
$
(B^\ast, \Theta, \mathbf{y})
$
denote the observed data defined in \eqref{def:problem_instance}.
At the $l$-th iteration of the PLS algorithm, let
$( \hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\xi}}^\top )^\top$
denote the coefficients of $\hat{\xi}$, defined in \eqref{def:maximizer_squared_empirical_cov},
which corresponds to the projected unit-norm direction that maximizes the squared empirical covariance.
These coefficients are obtained by solving the following generalized eigenvalue problem:
\begin{equation}	\label{eq: Regularized generalized rayleigh quotient equation}
	\left(
	\hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\xi}}^\top
	\right)^\top
	=
	\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
	\quad \text{subject to} \quad \boldsymbol{\xi}^\top B^\ast \boldsymbol{\xi} = 1,
\end{equation}
where
\begin{equation}\label{def:V_matrix}
V := \frac{1}{n^2} (B^\ast \Theta^\top \mathbf{y})(B^\ast \Theta^\top \mathbf{y})^\top \in \mathbb{R}^{(MK+p) \times (MK+p)},
\end{equation}

@@

		\begin{align}
			\begin{split}
				&\arg \max \limits_{\xi^* \in \mathbb{R}^{MK+p}} \; \frac{\xi^{*\top} \mathbf{V}^\ast \xi^*}{\xi^{*\top}(J^\ast+\Lambda \ddot{J}^\ast) \xi^*} , \quad \textit{or equivalently}, \\ 
				&\arg \max \limits_{\xi^* \in \mathbb{R}^{MK+p}} \;  \xi^{*\top} \mathbf{V}^\ast \xi^* \quad \textnormal{subject to}  \quad \xi^{*\top} (J^\ast+\Lambda \ddot{J}^\ast) \xi^*=1.
				\label{eq: Regularized generalized rayleigh quotient equation}
			\end{split}
		\end{align}
		We then express the smoothed component in terms of the basis functions as
		\begin{equation*}
			\widehat{\xi}[\mathbf{t}] = [\widehat{\psi}(\mathbf{t})^\top, \widehat{\boldsymbol{\theta}}^\top]^\top =  B^\ast[\mathbf{t}]^\top \widehat{\xi}^\ast.   
		\end{equation*}
		The $l$-th PLS score $\widehat{\rho}_{i} \equiv \widehat{\rho}_{il}$ can be obtained as 
		\begin{equation}\label{compute_pls_score}
			\widehat{\rho}_i = \langle \mathbf{W}_i, \widehat{\xi} \rangle_\mathcal{H}
		\end{equation}
	\end{proposition}
	The proof of Proposition \ref{Proposition: Regularized generalized rayleigh quotient} is provided in Appendix \ref{section:proof:eigenproblem_derivation}.
	Here, $\lambda_k$ $(k=1, \ldots, K)$ is a positive smoothing parameter that controls the trade-off between the goodness of fit and the amount of smoothness in $\psi^{(k)}$. Smaller $\lambda_k$ produces PLS components that fit more closely to the observed data, but selecting too small value may cause overfitting. The estimated PLS component reverts back to the unregularized version of Proposition \ref{proposition:eigen_noregul} when $\lambda_k=0$ for all $k$. On the other hand, larger $\lambda_k$ places more emphasis on smoothing $\psi^{(k)}$. In the limit $\lambda_k \rightarrow \infty$, each of the functions is forced to be of the form $\psi^{(k)}(t)=a+bt$ for some constants $a$ and $b$. In practice, $\{\lambda_k\}_{k=1}^K$ as well as the number of PLS components $L$ to be estimated can be chosen by cross-validation based on a certain predictive performance criterion (e.g., mean squared error).

@@

B^\ast  

@@

(B^\ast \Theta^\top \mathbf{y})(B^\ast \Theta^\top \mathbf{y})^\top 

@@

	\boldsymbol{\xi}^\top B^\ast \boldsymbol{\xi}.

>>>>

	\subsubsection{Finite-basis approximation of the predictors and regression coefficients}\label{sec: finite basis approximation}	
	Given a twice-differentiable basis $\{b_l(t)\}$ of $L^2([0,1])$, the functional predictors and the corresponding regression coefficients can be decomposed into
	\begin{equation*}
		X_{ij}(t) = \sum_{l=1}^\infty \theta_{ijl} \, b_l(t), \quad 
		\beta_j(t) = \sum_{l=1}^\infty \eta_{0,jl} \, b_l(t),
	\end{equation*}
	where $\theta_{ijl}$ and $\eta_{0,jl}$ are the coefficients of $X_{ij}(t)$ and $\beta_j(t)$ corresponding to the $l$th basis function $b_l(t)$, respectively.
Since functional predictors are observed at finitely many discrete time points in practice,
we use finite-dimensional projections obtained by the expansion truncating at $M$ terms.
We choose a moderately large value of $M$ (e.g., 15 or 20) to capture rich functional variation without requiring fine-tuning, as our method promotes smoothness through penalization (see Section~\ref{sec: Regularization of PLS Components}). We denote the truncated space by
	\begin{equation}
		\mathcal{B} := \operatorname{span} \bigl( b_1(t), \ldots, b_M(t) \bigr).
	\end{equation}
	 We denote the  predictors projected onto $\mathcal{B}$ and the resulting coefficients as
	\begin{equation*}
		\widetilde{X}_{ij}(t) = \sum_{l=1}^M \theta_{ijl} \, b_l(t), \quad 
		\tilde{\beta}_j(t) = \sum_{l=1}^M \eta_{0,jl} \, b_l(t).
	\end{equation*}
	The $i$th  hybrid predictor with projected functional parts, is a tuple, denoted as
	\begin{equation*}
		\widetilde{W}_i := 
		(
		X_{i1}, \ldots, X_{iK}, \boldsymbol{Z}_i
		).
	\end{equation*}
	We adopt the following notation for the coefficients:
	\begin{equation}
		\boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top \in \mathbb{R}^M,
		\quad
		\boldsymbol{\eta}_{0j} := (\eta_{0,j1}, \ldots, \eta_{0,jM})^\top  \in \mathbb{R}^M,
	\end{equation}
	and
	\begin{equation}
		\Theta_j := (\boldsymbol{\theta}_{1j}, \ldots, \boldsymbol{\theta}_{nj})^\top  \in \mathbb{R}^{n \times M}.
	\end{equation}

The observations in the hybrid PLS problem can be compactly represented by the tuple
\begin{equation}\label{def:problem_instance}
	(\mathbb{B}, \Theta, \mathbf{y})
	\in
	,
\end{equation}
defined as follows.
Let $B, B^{\prime\prime} \in \mathbb{R}^{M \times M}$ denote the Gram matrices of the basis functions and their second derivatives, respectively, with entries  
\begin{equation}
	B_{m, m'} := \int_0^1 b_m(t)\, b_{m'}(t)\, dt,
	\quad
	B^{\prime\prime}_{m, m'} := \int_0^1 b^{\prime\prime}_m(t)\, b^{\prime\prime}_{m'}(t)\, dt,
\end{equation}
for $m, m' = 1, \ldots, M$.

Define the $(MK + p) \times (MK + p)$ block-diagonal matrices
\[
\mathbb{B} := \operatorname{blkdiag}(B, \ldots, B, I_p)
\quad \text{and} \quad
\mathbb{B}^{\prime\prime} := \operatorname{blkdiag}(B^{\prime\prime}, \ldots, B^{\prime\prime}, I_p),
\]
and let $\Theta := (\Theta_1, \ldots, \Theta_K, Z)$ denote the combined design matrix. Finally, let $\mathbf{y} := (y_1, \ldots, y_n)^\top$ denote the observed response vector.

	\begin{remark}
While it is possible to use a different basis for each functional predictor, we adopt a common basis for all predictors to simplify the notation. However, the definition of $\mathbb{B}$ remains general enough to accommodate different bases across the $K$ predictors.
\end{remark}

@@

	\begin{proposition}\label{proposition:eigen_noregul}
Let
$
(\mathbb{B}, \Theta, \mathbf{y})
$
denote the observed data defined in \eqref{def:problem_instance}.
At the $l$-th iteration of the PLS algorithm, let
$( \hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\xi}}^\top )^\top$
denote the coefficients of $\hat{\xi}$, defined in \eqref{def:maximizer_squared_empirical_cov},
which corresponds to the projected unit-norm direction that maximizes the squared empirical covariance.
These coefficients are obtained by solving the following generalized eigenvalue problem:
\begin{equation}	\label{eq: Regularized generalized rayleigh quotient equation}
	\left(
	\hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\xi}}^\top
	\right)^\top
	=
	\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
	\quad \text{subject to} \quad \boldsymbol{\xi}^\top \mathbb{B} \boldsymbol{\xi} = 1,
\end{equation}
where
\begin{equation}\label{def:V_matrix}
V := \frac{1}{n^2} (\mathbb{B} \Theta^\top \mathbf{y})(\mathbb{B} \Theta^\top \mathbf{y})^\top \in \mathbb{R}^{(MK+p) \times (MK+p)},
\end{equation}

@@

		\begin{align}
			\begin{split}
				&\arg \max \limits_{\xi^* \in \mathbb{R}^{MK+p}} \; \frac{\xi^{*\top} \mathbf{V}^\ast \xi^*}{\xi^{*\top}(J^\ast+\Lambda \ddot{J}^\ast) \xi^*} , \quad \textit{or equivalently}, \\ 
				&\arg \max \limits_{\xi^* \in \mathbb{R}^{MK+p}} \;  \xi^{*\top} \mathbf{V}^\ast \xi^* \quad \textnormal{subject to}  \quad \xi^{*\top} (J^\ast+\Lambda \ddot{J}^\ast) \xi^*=1.
				\label{eq: Regularized generalized rayleigh quotient equation}
			\end{split}
		\end{align}
		We then express the smoothed component in terms of the basis functions as
		\begin{equation*}
			\widehat{\xi}[\mathbf{t}] = [\widehat{\psi}(\mathbf{t})^\top, \widehat{\boldsymbol{\theta}}^\top]^\top =  \mathbb{B}[\mathbf{t}]^\top \widehat{\xi}^\ast.   
		\end{equation*}
		The $l$-th PLS score $\widehat{\rho}_{i} \equiv \widehat{\rho}_{il}$ can be obtained as 
		\begin{equation}\label{compute_pls_score}
			\widehat{\rho}_i = \langle \mathbf{W}_i, \widehat{\xi} \rangle_\mathcal{H}
		\end{equation}
	\end{proposition}
	The proof of Proposition \ref{Proposition: Regularized generalized rayleigh quotient} is provided in Appendix \ref{section:proof:eigenproblem_derivation}.
	Here, $\lambda_k$ $(k=1, \ldots, K)$ is a positive smoothing parameter that controls the trade-off between the goodness of fit and the amount of smoothness in $\psi^{(k)}$. Smaller $\lambda_k$ produces PLS components that fit more closely to the observed data, but selecting too small value may cause overfitting. The estimated PLS component reverts back to the unregularized version of Proposition \ref{proposition:eigen_noregul} when $\lambda_k=0$ for all $k$. On the other hand, larger $\lambda_k$ places more emphasis on smoothing $\psi^{(k)}$. In the limit $\lambda_k \rightarrow \infty$, each of the functions is forced to be of the form $\psi^{(k)}(t)=a+bt$ for some constants $a$ and $b$. In practice, $\{\lambda_k\}_{k=1}^K$ as well as the number of PLS components $L$ to be estimated can be chosen by cross-validation based on a certain predictive performance criterion (e.g., mean squared error).

@@

\mathbb{B}  

@@

(\mathbb{B} \Theta^\top \mathbf{y})(\mathbb{B} \Theta^\top \mathbf{y})^\top 

@@

	\boldsymbol{\xi}^\top \mathbb{B} \boldsymbol{\xi}.
